Topological Equivariant Artist Model for Visualization Library Architecture

### 2023-11-03 paper/main.tex
put intro back together

 In this paper, we propose a methodology for describing structure, verifying structure preservation, and specifying the conditions for constructing a structure preserving map between data and graphics using topology and category theory.

 We use category theory because it is the math of structures and their relations and therefore provides a robust framework for defining structure and expressing the conditions under which structure would be preserved.

 We encode the structure of the data and graphics in a uniform generalizable manner ; doing so allows us to discuss everything from lines to complex ND spaces with the same interface.

We use category theory and algebraic structures because they can be directly translated into software design specifications and our goal is to improve visualization tools. The algebraic structures can be translated into analogous programmatic types, while category theory facilitates developing a functional design framework that describes function composition in terms of functional algebraic operations. The typing system and composition inherent in a functional design encourage library developers to build complex visualization components from simple verifiable parts that have few side effects\cite{huHowFunctionalProgramming2015, hughesWhyFunctionalProgramming1989}. These categorically specified components could be built as a standalone library and integrated into existing libraries because they are inherently self-contained. Furthermore we see the application of these ideas in low level visualization library design as our motivation, and we hope these ideas will help shape the reorganization of critical data visualization libraries, such as Matplotlib. The contribution of this paper is a mathematical framework for describing structure preserving visualization components in a manner that is generalizable to different types of inherent structure. This framework also provides guidance for the construction and testing of structure preserving visualization library components.

>>>>

We propose a methodology for describing structure, verifying structure preservation, and specifying the conditions for constructing a structure preserving map between data and graphics using topology and category theory. Using category theory and algebraic structures, we can encode the structure of data and graphics in a uniform manner. This uniformity allows us to develop a generalizable framework for specifying the conditions under which structure is preserved. For example, our methodology can be used to describe how a table and scatter plot and heatmap are different representations of the same data and track an observation from a datacube as a point along a time series and on a map and in a network. We developed a generalizable model because our goal is to improve data visualization software. The algebraic structures can be translated into analogous programmatic types, while category theory facilitates developing a functional design framework. Strong typing and function composition enable library developers to build complex components from simpler verifiable parts \cite{huHowFunctionalProgramming2015, hughesWhyFunctionalProgramming1989}. These  components can be built as a standalone library and integrated into existing libraries and we hope these ideas will help shape the reorganization of critical data visualization libraries, such as Matplotlib.

The contribution of this paper is a mathematical framework for describing structure preserving visualization components in a manner that is generalizable to different types of inherent structure. This framework also provides guidance for the construction and testing of structure preserving visualization library components.
